Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-06 aistats.tex
Clean up open problem

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\natp)$ is
\begin{align}
    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is introduced above.
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
Then the joint log-likelihood of $\cD =(X_1,\dots,X_n)$ and $\natp$ is
\begin{multline}
    -\log p(\cD|\natp)p(\natp) 
    = (n_0+n) \logpart (\natp) \\
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
    \label{eq:joint_likelihood}
\end{multline}
where $T_i=T(X_i)$.
This means the posterior given $\cD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.

\paragraph{Maximum A Posteriori (MAP).}
Minimizing~\eqref{eq:joint_likelihood} over $\natp$ yields the MAP estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
\end{align}
If they can be satisfied, the MAP verifies the first oder optimality conditions
\begin{align}
    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
These rules are known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.

@@

\fdk{
Might want to note that a perhaps ``typical'' view of parameter estimation 
works with norms, e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ (for some norm and some power), 
or more generally using distances rather than divergences as they are symmetric and have nicer properties. 
But as the MLE is the minimizer of the log-likelihood, 
using the log-likelihood (and by extension the KL divergence) as a measure of risk is natural, if inconvenient.
}

For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{multline}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{multline}
For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments 
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp)
}
\end{align}
Our question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\cD]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \\
	\label{eq:bregmanMAP}
	\expect[\cD]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} 
\end{align}
where the outer expectation is on the dataset $\cD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.

>>>>

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\natp)$ is
\begin{align}
    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is introduced above.
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\natp$ is
\begin{multline}
    -\log p(\mD|\natp)p(\natp) 
    = (n_0+n) \logpart (\natp) \\
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
    \label{eq:joint_likelihood}
\end{multline}
where $T_i=T(X_i)$.
This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.

\paragraph{Maximum A Posteriori (MAP).}
Minimizing~\eqref{eq:joint_likelihood} over $\natp$ yields the MAP estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(\mD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
\end{align}
If they can be satisfied, the MAP verifies the first oder optimality conditions
\begin{align}
    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
These rules are known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.

@@

Let $\natp^* := \argmin \expect[X\sim\cD]{-\log p(X|\natp)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{multline}
    \expect[X\sim \cD]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{multline}
Note that $\cD$ does not have to be part of the exponential family.
For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments 
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp)
}
\end{align}

Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \TODO{cite} 
Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \TODO{cite} 
A contrario the KL is not smooth, and it can be infinite in misspecified settings.
But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.

Our question is: how does the KL behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \mu_0 + \smallsum_i T_i}{n_0+n} \right )} 
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.
